---
title: "A549-HBE-ALI comparison analysis"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# for set theory
library(ggVennDiagram)
library(ComplexUpset)
library(eulerr)

# misc tools
library(patchwork)
source("modules/helpers.R")

#tidyverse
library(tidyverse)
```

<!-- All set up!!! Repo is at https://github.com/axg610/a549-hbe-ali-comparison. -->

```{r load data}
dea <- read_tsv("data/raw/A549vsPrimary_univariateDEA.txt")
tpm <- read_tsv("data/raw/A549vsPrimary_tpm.txt")
multivar <- read_tsv("data/raw/A549vsPrimary_multivariateDEA.txt")
```

## Alignment and DEA

**Raw datasets:** A549, primary hBECs (submersion culture, HBE), and primary hBECs (air-liquid interface culture, ALI) treated with IL1B, glucocorticoid, or both at 6h. This is actually three experiments at different timepoints, so for now we use their shared timepoint (6 h post-treatment).

Alignment performed with kallisto version 0.46.1 and normalization/DEA performed with sleuth version something on the cluster. Everything documented in `kallisto-sleuth.md`. I used individual sleuth runs for each cell type, modeled as `~treatment`. Effect sizes ("log2fold") and FDR are from Wald test using unstimulated (NS) samples as baseline. I also modeled everything together under model `~treatment*condition` for interaction testing. Finally, raw tpm values were extracted.

```{r data previews}
knitr::kable(
  dea %>% filter(Gene == "NFKBIA"),
  caption = "Example univariate DEA data (~treatment)", 
  align = "c"
)

knitr::kable(
  multivar %>% filter(Gene %in% c("NFKBIA", "TNFAIP3", "IRAK3")),
  caption = "Example multivariate DEA data (~treatment)", 
  align = "c"
)

knitr::kable(
  tpm %>% filter(Gene == "NFKBIA") %>% .[1:20,],
  caption = "Example tpm data, first 20 rows for representative gene", 
  align = "c"
)
```

## IN PROGRESS: Understanding individual variation in gene expression

Previously, we saw that **only ~17% of the IL1B-regulated gene signature in A549s is shared with at least one primary celltype**. See Venn diagrams below for DEGs (log2fold <= -1 or >= 1 & FDR < 0.05) in each cell type (panel A), up-DEGs (panel B), and down-DEGs (panel C). When looking for "regulation" in general, A549s look very different from primary cells. Splitting into up- and down-regulated groups reveals that most of this difference is attributable to downregulated genes. i.e., A549s are a reasonable model for IL1B-responsive gene induction but not suppression in primary cells. If we just look at upregulated genes, **~33% of the IL1B-regulated gene signature in A549s is shared with at least one primary celltype**.

**TODO: what % of the IL1B-regulated signature in A549s is shared with at least one donor?**

```{r old venns, fig.height=4, fig.width=5}
source("modules/fig-old-venns.R")
cleanup_objects()
```

One possible reason for differences: A549s are derived from one individual, so four sequencing replicates may have low variability + higher significance. Each sequencing replicate for primary cells is a different individual, so higher variability + lower significance. Ignoring FDR entirely (i.e., only using the abs(log2fold) >= 1 threshold), A549 and primary cells actually have similarly sized regulated gene pools (panel A). In A549s, variation in tpm-derived log2fold values tend to be lower (panel B) and a greater proportion of the rep pool meets log2fold thresholds (panel C).

*Fixed: problem 040626, discrepancy between Venn and spread gene counts.*

```{r calculate per-donor metrics}
source("modules/per-donor-set-theory.R")
cleanup_objects()
```

```{r basic per-donor threshold metrics, fig.height=3, fig.width=8}





cleanup_objects()
```

**Where does each primary donor sit in relation to A549?** For each primary cell donor, check overlap of log2fold-threshold-meeting genes in that donor **(top circle)** vs A549 cells **(bottom circle)**. Plots are also tagged with Jaccard index **(J)** and total number of genes **(n, union size).** This analysis was performed on regulated genes in general (panel A), IL1B-upregulated genes (panel B), and IL1B-downregulated genes (panel C)

*"Fixed" colors on Venns. This package doesn't let you specify colors for different groups, so I set everything to white instead.*

```{r pairwise venns vs a549, fig.width=8}



# write overlaps to the main per-donor object



# plots

plot_pairwise_venns <- function(
    inputOverlaps, 
    referenceGroup = a549_genes,
    plotTitle = "title here"){
  
  # take an overlaps dataframe from above and create an array of pairwise
  # venn digrams using ggVennDiagram.
  
  # must specify the reference group (the bottom venn lobe)
  
  plots <- inputOverlaps %>%
    mutate(
    p = map2(
      genes,
      title,
      ~ ggVennDiagram(
        list(
          " " = referenceGroup,
          "  " = .x
        ),
        label_alpha = 0,
        label_size = 3
      ) +
        ggtitle(.y) +
        scale_fill_gradient(low = "white", high = "white") +
        theme(
          legend.position = "none",
          plot.title = element_text(size = 10, hjust = 0.5)
        )
    )
  )
  
  print(
    wrap_plots(plots$p, ncol = 6) +
      plot_annotation(title = plotTitle)
    )
  
}

plot_pairwise_venns(
  overlaps,         
  a549_genes, 
  "(A) Pairwise Venns for IL1B-reg genes: abs(log2fold) >= 1, donor (top) vs A549 (bottom)"
  )

plot_pairwise_venns(
  overlaps_upreg,   
  a549_genes_upreg, 
  "(B) Pairwise Venns for IL1B-upreg genes: log2fold >= 1, donor (top) vs A549 (bottom)"
  )

plot_pairwise_venns(
  overlaps_downreg, 
  a549_genes_downreg, 
  "(C) Pairwise Venns for IL1B-downreg genes: log2fold <= -1, donor (top) vs A549 (bottom)"
  )

cleanup_objects()
```

Some summary metrics for the pairwise Venns:

- Union = everything in the A549 lobe, the donor lobe, and the intersect
- Overlap = genes expressed in both A549 and the donor
- Jaccard = $\frac{\text{overlap}}{\text{union}}$, high = big overlap, low = little to no overlap

Each dot is one donor-to-A549 comparison. We see that overlaps are bigger when focusing on upregulated genes, and smaller when focusing on downregulated genes. Again showing that most commonality between cell types sits in the upregulated gene fraction. T-test between $Jaccard_{\text{upreg}}$ and $Jaccard_{\text{downreg}}$, the bottom right panel, shows P=0.0001169.

```{r pairwise venns summary, fig.height=4, fig.width=4}
df <- per_donor_data[["overlaps_summary"]] %>%
  filter(name != "A549 mean") %>%
  select(criterion, name, union, overlap, jaccard) %>%
  pivot_longer(cols = c(union, overlap, jaccard), names_to = "stat", values_to = "value") %>%
  mutate(
    group = if_else(criterion == "reg", "all regulated genes", "split by direction"),
    stat = factor(stat, levels = c("union", "overlap", "jaccard"))
    )

ggplot(df,
       aes(x = criterion, y = value)
       ) +
  facet_grid(stat~group, scales = "free") +
  stat_summary(geom = "bar", fun = "mean", width = 0.5, color = "black") +
  # stat_summary(geom = "errorbar", fun.data = "mean_se", width = 0.3) +
  geom_jitter(width = 0.05) +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid = element_blank()
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25))) +
  labs(y = NULL, x = NULL)

# t.test(
#   per_donor_data[["overlaps_summary"]] %>% 
#     filter(criterion == "upreg", name != "A549 mean") %>% 
#     pull(jaccard),
#     per_donor_data[["overlaps_summary"]] %>% 
#     filter(criterion == "downreg", name != "A549 mean") %>% 
#     pull(jaccard)
# )
```

**Where do primary donors sit in relation to each other?** This is technically a 10-way comparison so Venns no longer work -- use an UpSet plot instead. Panel A is the intersect between *all donors and A549*, B is the intersect between *donors only*. Notice 70 genes shared between all donors (B), where 54 of these are also shared with A549 (A).

```{r calculate matrices for upset plots}


  
# write logical matrices to the main per-donor object



cleanup_objects()
```


```{r upset plots between donors, fig.height=8, fig.width=7}

# plots

p1 <- plot_upset(logicalMatrix = per_donor_data[["logical_matrix"]], plotTitle = "A")
p2 <- plot_upset(per_donor_data[["logical_matrix_a5Exclude"]], plotTitle = "B")

wrap_plots(p1, p2, nrow = 2)

cleanup_objects()
```

54 "universally" shared genes from above:

```{r}
per_donor_data[["logical_matrix"]] %>%
  filter(if_all(everything(), ~ .x == 1)) %>%
  rownames()
```

70 donor-shared genes from above:

```{r}
per_donor_data[["logical_matrix"]] %>%
  select(-`A549 mean`) %>%
  filter(if_all(everything(), ~ .x == 1)) %>%
  rownames()
```

16 (70-54) donor-only genes:

```{r}
setdiff(
  per_donor_data[["logical_matrix"]] %>%
    select(-`A549 mean`) %>%
    filter(if_all(everything(), ~ .x == 1)) %>%
    rownames(),
  per_donor_data[["logical_matrix"]] %>%
    filter(if_all(everything(), ~ .x == 1)) %>%
    rownames()
)
```

Now split the visualization down from all regulated genes to upregulated and downregulated genes (see below). Notice how 50 of the 54 "universally shared" genes present in the upregulated fraction (panel A), and 70 of 70 "donor-shared" genes present in the upregulated fraction (panel C). However, DEGs unique to each donor still dominate over DEGs shared between donors.

```{r, fig.height=8, fig.width=9}

p1 <- plot_upset(logicalMatrix = per_donor_data[["logical_matrix_upreg"]], plotTitle = "(A) IL1B-upregulated")
p2 <- plot_upset(logicalMatrix = per_donor_data[["logical_matrix_downreg"]], plotTitle = "(B) IL1B-downregulated")


p3 <- plot_upset(per_donor_data[["logical_matrix_a5Exclude_upreg"]], plotTitle = "(C) IL1B-upreg, primary only")
p4 <- plot_upset(per_donor_data[["logical_matrix_a5Exclude_downreg"]], plotTitle = "(D) IL1B-downreg, primary only")

wrap_plots(p1, p2, p3, p4, nrow = 2)

cleanup_objects()
```

<!-- *TODO: visualize as a rough Euler plot* -->

```{r}
# per_donor_data[["logical_matrix_upreg"]] %>% 
#   plot_euler(plotTitle = "IL1B-upreg genes (6h, some small intersections ignored)")
# 
# test <- per_donor_data[["logical_matrix_upreg"]] %>%
#   filter(`A549 mean` == 1) %>%
#   plot_euler(plotTitle = "IL1B-upreg genes at 6h (only intersections involving A549)") 
# 
# test
```





