---
title: "A549-HBE-ALI comparison analysis"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# for functional annotation
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)

# for set theory
library(ggVennDiagram)
library(ComplexUpset)
library(eulerr)

# heatmaps
library(ComplexHeatmap)
library(pheatmap)

# misc tools
library(patchwork)
library(ggrepel)
source("modules/helpers.R")

#tidyverse
library(tidyverse)
```

Repo at https://github.com/axg610/a549-hbe-ali-comparison

We can edit in PowerPoint/apply a consistent theme to all the figures at the very end

# **1. Alignment and DEA**

**Raw datasets:** A549, primary hBECs (submersion culture, HBE), and primary hBECs (air-liquid interface culture, ALI) treated with IL1B, glucocorticoid, or both at 6h. This is actually three experiments at different timepoints, so for now we use their shared timepoint (6 h post-treatment).

Alignment performed with kallisto version 0.46.1 and normalization/DEA performed with sleuth version ??? on the cluster. Everything documented in `kallisto-sleuth.md`. I used individual sleuth runs for each cell type, modeled as `~treatment`. Effect sizes ("log2fold") and FDR are from Wald test using unstimulated (NS) samples as baseline. I also modeled everything together under model `~treatment*condition` for interaction testing. Raw tpm values were also extracted.

```{r load data, cache=T}
dea <- read_tsv("data/raw/A549vsPrimary_univariateDEA.txt")
tpm <- read_tsv("data/raw/A549vsPrimary_tpm.txt")
multivar <- read_tsv("data/raw/A549vsPrimary_multivariateDEA.txt")
```

```{r data previews, cache=T}
# knitr::kable(
#   dea %>% filter(Gene == "NFKBIA"),
#   caption = "Example univariate DEA data (~treatment)", 
#   align = "c"
# )
# 
# knitr::kable(
#   multivar %>% filter(Gene %in% c("NFKBIA", "TNFAIP3", "IRAK3")),
#   caption = "Example multivariate DEA data (~treatment*condition)", 
#   align = "c"
# )
# 
# knitr::kable(
#   tpm %>% filter(Gene == "NFKBIA") %>% .[1:20,],
#   caption = "Example tpm data, first 20 rows for representative gene", 
#   align = "c"
# )
```

# **2. Story notes**

Given that IL1B-responsive gene expression patterns in A549s cells follow ... kinetics, carry ... functional signatures, we next want to compare the transcriptomic signatures of A549 cells and primary cells at 6 h (time of most regulation)... With the ultimate aim of understanding the extent to which A549 cells recapitulate donor responses to IL1B.

## 2.1. Characterizing variability in IL1B response at the donor level

A549 cells are derived from one individual, where each of the four replicates reflects one donor. Primary cells are derived from different individuals, where each replicate corresponds to a different donor. Thus, donor-specific regulatory events will be penalized in primary cell data and emphasized in A549 data. Before proceeding with differential expression analysis, it is important to assess IL1B-responsive transcription at the individual level.

**Primary cell donors are more variable than A549 cell replicates.** Counts and variability for regulated genes (abs(log2fold) > 1)  were assessed across sequencing replicates. **(A)** A549 cells have more or less the same number of "regulated" genes across replicates, and **(B)** lower variability in log2fold across replicates (*do stats on this?*). **(C)** A549 replicates are more concordant in which genes they regulate while individuals donors are more variable. **(D)** When visualizing the union of all regulated genes (14621) across replicates as a heatmap, A549 cells tend to show consistent responses across replicates while primary cells show blocks of up- or down-regulation on a per-donor basis.


```{r calculate per-donor metrics, cache=T}
source("modules/per-donor-set-theory.R")
source("modules/compute-heavy-heatmaps.R")
cleanup_objects()
```

```{r generate per-donor figs, cache=T, include=F}
source("modules/fig-per-donor-threshold-metrics.R")
```

```{r plot per-donor figs 1, fig.height=3.5, fig.width=8}
temp_figs[[1]]
```

```{r plot per-donors figs 2, fig.height=7, fig.width=5, out.width="50%", fig.show='hold'}
temp_figs[[2]]
temp_figs[[3]]
cleanup_objects()
```

**Donors show more uniquely regulated genes than shared genes.** To understand individual variation, A549 cell replicates were averaged to represent one donor. Then overlaps between the A549 donor and all other primary donors were visualized as an UpSet plot. **(A)** Compared to genes regulated in only one donor (left side, first 11 bars), there are much fewer genes shared between donors (right side) *(do a manual count?)*. Splitting the visualization into **(B)** upregulated and **(C)** downregulated genes reveals that A549s have "normal" IL1B-upregulated signatures (middle-of-the-pack in terms of unique upreg genes), but abberant IL1B-downregulated signatures (has the most downregulated genes). Additionally, the 54 completely conserved genes are exclusively upregulated. This suggests that the IL1B-induced response is more core/conserved while the downregulated response is more context/individual-specific.

```{r plot upsets between donors, fig.height=8, fig.width=8}
source("modules/fig-per-donor-upset.R")
cleanup_objects()
```

**Transcriptomic overlaps between each donor and A549 cells are small.** To assess similarities between each primary cell donor and the A549 mean, regulated genes were compared between each primary cell donor and A549 cells. **(A)** Only a small percentage (<15%) of each donor's regulation pool is recapitulated by A549 cells. This is less of a problem for IL1B-induced genes **(B)** and more of a problem for IL1B-downregulated genes **(C)**. J = Jaccard index, n = total number of genes.

```{r plot pairwise venns vs a549, fig.width=8}
source("modules/fig-per-donor-pairwise-venns.R")
cleanup_objects()
```

Plotting each donor-to-A549 comparison, we see that overlaps tend to be bigger for upregulated genes and smaller for downregulated. T-test for $Jaccard_{\text{upreg}}$ vs $Jaccard_{\text{downreg}}$ (bottom right panel) returns P=0.0001169 (***).

```{r plot pairwise venns summary, fig.height=4, fig.width=4}
source("modules/fig-per-donor-pairwise-venns-summary.R")
cleanup_objects()
```

## [[IN PROGRESS]] 2.2. Functional consequences of between-donor variability

**Inflammatory programs are stably regulated across donors.** To understand which transcriptional programs are highly variable vs conserved across donors, the union of all IL1B-induced genes in each cell type were ranked by -SD across all donors. Thus, top ranking genes (green) are consistently/stably regulated by IL1B in all donors, while bottom ranking genes (purple) are highly variable between donors.

```{r run gene ontology on variable vs stable genes, cache=T}

go_results <- list()

go_results[["variable_vs_stable"]][["df"]] <- data.frame()

for(ct in c("A549", "ALI", "HBE")){
  
  # rank all "regulated" genes by SD between donors
  
  ranked <- tpm %>%
    filter(
      celltype == ct,
      treatment == "IL1B"
    ) %>%
    group_by(Gene) %>%
    filter(any(log2fold >= 1)) %>%
    summarize(sd = -sd(log2fold), .groups = "drop") %>%
    arrange(desc(sd)) %>%
    deframe()
  
  # ranked list of all genes
  all_genes <- names(ranked)
  
  # calculate and cut the top/bottom 10%
  n <- length(all_genes)
  cut <- floor(0.10 * n)
  top_genes <- all_genes[1:cut]
  bottom_genes <- all_genes[(n - cut + 1):n]
  
  # run and extract top results, enrichGO on biological process
  
  top_go_bp <- enrichGO(
    gene          = top_genes,
    universe      = all_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )
  
  top_go_result <- top_go_bp@result %>%
    mutate(
      celltype = ct,
      treatment = "IL1B",
      group = "top 10%",
      .before = 1
    )
  
  rownames(top_go_result) <- NULL
  
  # run and extract bottom results, enrichGO on biological process
  
  bottom_go_bp <- enrichGO(
    gene          = bottom_genes,
    universe      = all_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )
  
  bottom_go_result <- bottom_go_bp@result %>%
    mutate(
      celltype = ct,
      treatment = "IL1B",
      group = "bottom 10%",
      .before = 1
    )
  
  rownames(bottom_go_result) <- NULL
  
  # save dataframes to main object
  
  go_results[["variable_vs_stable"]][["df"]] <- rbind(
    go_results[["variable_vs_stable"]][["df"]], 
    top_go_result
    )
  
  go_results[["variable_vs_stable"]][["df"]] <- rbind(
    go_results[["variable_vs_stable"]][["df"]], 
    bottom_go_result
  )
  
  # save genelists to main object
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_top10percentInduced")
  ]] <- top_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_bottom10percentInduced")
  ]] <- bottom_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_allInduced")
  ]] <- all_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_allInduced_sd")
  ]] <- ranked
  
}

cleanup_objects()
```

```{r plot heatmap of sd rankings, fig.height=4, fig.width=4, out.width='50%', fig.show='hold'}

for(ct in c("ALI", "HBE")){
  
  # get matrix and genelists
  query <- paste0(ct, "_IL1B_allInduced")
  topquery <- paste0(ct, "_IL1B_top10percentInduced")
  bottomquery <- paste0(ct, "_IL1B_bottom10percentInduced")
  
  genelist <- go_results[["variable_vs_stable"]][[query]]
  top_genes <- go_results[["variable_vs_stable"]][[topquery]]
  bottom_genes <- go_results[["variable_vs_stable"]][[bottomquery]]
  
  mat <- tpm %>%
    filter(
      treatment == "IL1B",
      Gene %in% genelist,
      celltype == ct
    ) %>%
    select(Gene, rep, log2fold) %>%
    pivot_wider(values_from = "log2fold", names_from = "rep") %>%
    mutate(Gene = factor(Gene, levels = genelist)) %>%
    arrange(Gene) %>%
    column_to_rownames("Gene") %>%
    as.matrix()
  
  # set colors
  colors <- generic_l2f_heatmap_colors(mat)
  
  # assign groups by membership
  group <- case_when(
    rownames(mat) %in% top_genes ~ "top 10% -SD (most stable)",
    rownames(mat) %in% bottom_genes ~ "bottom 10% -SD (most variable)",
    TRUE ~ "middle"
  )
  
  group <- factor(
    group, 
    levels = c(
      "top 10% -SD (most stable)", 
      "middle", 
      "bottom 10% -SD (most variable)"
      )
    )
  
  # annotation
  row_ha <- rowAnnotation(
    group = group,
    col = list(
      group = c(
        "top 10% -SD (most stable)" = "aquamarine3",
        middle = "grey77",
        "bottom 10% -SD (most variable)" = "orchid3"
      )
    ),
    show_annotation_name = FALSE,
    border = TRUE
  )
  
  # draw hm
  hm <- Heatmap(
    mat,
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    show_column_dend = FALSE,
    show_row_names = FALSE,
    col = colors,
    name = "log2fold",
    column_title = ct,
    height = unit(3, "inches"),
    width = unit(1, "inches"),
    left_annotation = row_ha,
    border = TRUE
  )
  
  print(hm)
  
}

cleanup_objects()


```

The two extremes of these rankings were taken for gene ontology enrichment analysis (GO_BP_DIRECT), and then the top 10 terms in each case were plotted. Terms with less than 10 genes were excluded to prevent poorly defined programs from dominated the ranking. **(C)** In both ALI and HBE cells, donors show stable induction of inflammatory and apoptotic programs. **(D)** Both ALI and HBE cells are more variable in expression of metabolic programs. HBE cells showed some enrichment for inflammatory responses. However, none of these enrichments reached significance.

```{r plot variable vs stable GO, fig.width=5, fig.height=7, out.width='50%', fig.show='hold'}

# function to prepare dataframes

prep_go <- function(df, ct, grp) {
  
  df %>%
    filter(celltype == ct, group == grp, Count >= 10) %>%
    slice_max(FoldEnrichment, n = 10, with_ties = FALSE) %>%
    mutate(
      Description = str_wrap(str_trunc(Description, 60), 35),
      Description = factor(Description, levels = rev(Description))
    )
  
}



# function to generate the base plot

base_plot <- function(
    df, 
    title,
    size_limits = c(0, 30),
    x_limits = c(0, 10),
    fill_limits = c(0, 4)) {
  ggplot(
    df,
    aes(
      x = FoldEnrichment,
      y = Description,
      fill = -log10(p.adjust),
      size = Count
    )
  ) +
    geom_point(shape = 21) +
    scale_size(
      limits = size_limits, range = c(0, 10)
    ) +
    # scale_fill_gradient(
    #   low = "white", high = "black",
    #   limits = fill_limits
    # ) +
    viridis::scale_fill_viridis(
      option = "viridis",
      limits = fill_limits
    ) +
    scale_x_continuous(
      limits = x_limits,
      expand = expansion(mult = 0.15)
    ) +
    labs(
      x = "fold enrichment", 
      y = title,
      size = "count",
      fill = expression(log[10] * "FDR")
    ) +
    theme_minimal()
}


# ==============================================================================


# show top 10% most stable genes in HBEs

p1 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "HBE", 
  grp = "top 10%"
  ) %>%
  base_plot(
    title    = "HBE", 
    x_limits = c(2.3, 4.0)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )

# show top 10% most stable genes in ALIs
p2 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "ALI", 
  grp = "top 10%") %>%
  base_plot(
    title    = "ALI", 
    x_limits = c(2.3, 4.0)
  ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank()
  )


# ==============================================================================


# show top 10% most variable genes in HBEs

p3 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "HBE", 
  grp = "bottom 10%"
  ) %>%
  base_plot(
    title    = "HBE", 
    x_limits = c(1, 1.9)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )

p4 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "ALI", 
  grp = "bottom 10%"
  ) %>%
  base_plot(
    title    = "ALI", 
    x_limits = c(1, 1.9)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank()
  )


# ==============================================================================

# assemble plot

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "A. most stably induced functions")

(p3 / p4) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "B. most variably induced functions")

cleanup_objects()
```








































# **3. Old or orphaned code (mostly junk)**

```{text, echo=T}

<!-- ## 2.?. IL1B responses seem to differ substantially between A549 and primary cells -->

<!-- **(A)** For genes deemed "regulated" by DEA (log2fold <= -1 or >= 1 & FDR < 0.05), only ~17% of the IL1B-regulated transcriptome is shared with at least one primary celltype. This figure increases to ~33% for IL1B-induced genes **(B)** and decreases to ~4% for IL1B-downregulated genes **(C)**. Overlaps remain robust when calling regulated genes with fold criteria only (~23% shared, **D**), or FDR only (~19% shared, **E**). -->

<!-- From a DEA perspective (all donors pooled together), IL1B-regulated transcriptomes in A549 and primary cells show low concordance. Start to see hints that the problem is worse for downregulated genes than it is for upregulated. -->

<!-- ```{r old venns, out.width="33%", fig.show='hold', cache=T} -->
<!-- source("modules/fig-old-venns.R") -->
<!-- cleanup_objects() -->
<!-- ``` -->





<!-- ## 2.?. A549s effectively capture what "could" happen across donors -->

<!-- ## 2.?. All celltypes share a small, conserved, "core" program -->

<!-- These genes have ... functional signature, follow ... kinetic, are ... NFkappaB dependent (and should be the focus of future studies?) -->







<!-- ## Understanding individual variation in gene expression -->

<!-- ### Donors vs A549 -->

<!-- // something used to be here // -->

<!-- **Where do primary donors and A549s sit in relation to each other?** -->

<!-- This is technically a 10-way comparison so Venns no longer work -- use an UpSet plot instead. Panel A is the intersect between *all donors and A549*, B is the intersect between *donors only*. Notice 70 genes shared between all donors (B), where 54 of these are also shared with A549 (A). -->

<!-- Now split the visualization down from all regulated genes to upregulated and downregulated genes (see below). Notice how 50 of the 54 "universally shared" genes present in the upregulated fraction (panel C), and 70 of 70 "donor-shared" genes present in the upregulated fraction (panel E). However, DEGs unique to each donor still dominate over DEGs shared between donors. -->


<!-- // something used to be here // -->


<!-- 54 "universally" shared genes from above: -->

<!-- ```{r} -->
<!-- per_donor_data[["logical_matrix"]] %>% -->
<!--   filter(if_all(everything(), ~ .x == 1)) %>% -->
<!--   rownames() -->
<!-- ``` -->

<!-- 70 donor-shared genes from above: -->

<!-- ```{r} -->
<!-- per_donor_data[["logical_matrix"]] %>% -->
<!--   select(-`A549 mean`) %>% -->
<!--   filter(if_all(everything(), ~ .x == 1)) %>% -->
<!--   rownames() -->
<!-- ``` -->

<!-- 16 (70-54) donor-only genes: -->

<!-- ```{r} -->
<!-- setdiff( -->
<!--   per_donor_data[["logical_matrix"]] %>% -->
<!--     select(-`A549 mean`) %>% -->
<!--     filter(if_all(everything(), ~ .x == 1)) %>% -->
<!--     rownames(), -->
<!--   per_donor_data[["logical_matrix"]] %>% -->
<!--     filter(if_all(everything(), ~ .x == 1)) %>% -->
<!--     rownames() -->
<!-- ) -->
<!-- ``` -->

<!-- ### Donors vs donors -->

<!-- **"Toy" example:** *at 5+ sets -- Venn impossible, Euler misleading, UpSet pretty good* -->

<!-- Starting with IL1B-upregulated genes in ALIs, I wanted to quantify how much of this signature is shared with A549s. The Euler diagram (panel A) suggested that only ~20% of genes were shared between A549s and at least one ALI donor -- which was surprising. Even more surprisingly, most donors did not have connections with A549s at all... -->

<!-- However, when we look at the full UpSet plot (panel B) we find that another ~20% of A549-donor-shared genes exist in small intersections involving one or two donors, which the Euler diagram cannot represent. A manual sum reveals that **A549s actually share 1488/2859 (52%) of IL1B-upregulated genes with at least one ALI donor.** -->


<!-- ```{r, out.width="50%", fig.show='hold'} -->
<!-- test <- per_donor_data[["logical_matrix"]] %>% -->
<!--   select(contains("A549"), contains("ALI")) %>% -->
<!--   filter(!if_all(everything(), ~ .x == 0)) -->

<!-- plot_euler(test, shapeType = "circle", plotTitle = "(A) a misleading Euler diagram") -->

<!-- plot_upset(test, minSize = 1, plotTitle = "(B) a less misleading UpSet plot") -->

<!-- cleanup_objects() -->
<!-- ``` -->

<!-- After writing and testing R code that could recreate the manually-calculated "52%" number from above, I scaled the analysis to ask; **for each donor, how much of its IL1B-regulated gene signature is shared with at least one other donor?** -->

<!-- (i) A549s have similar # of regulated genes when compared to donors (top left plot) -- aligns with previous observations -->

<!-- (ii) This holds true for up **and** downregulated genes (top right plot) -- ??? split the barplot from section 2.1. and see if this is correct??? -->

<!-- (iii) Given any donor, ~90% of its regulated genes are shared with at least one other donor (bottom left plot). A549s are an outlier at ~70%, but this is still pretty good. Basically means that ~70% of the transcriptome we study in A549s represents at least one of our donors. If we sequence more donors, the percentage could increase. -->

<!-- (iv) Low sharedness by A549s is mostly attributed to downregulated genes and not upregulated genes (bottom right plot). -->

<!-- ```{r, fig.height=4, fig.width=4} -->
<!-- # # confirm the manual 52% figure: -->
<!-- # -->
<!-- # df <- per_donor_data[["logical_matrix"]] %>% -->
<!-- #   select(contains("A549"), contains("ALI")) %>% -->
<!-- #   rownames_to_column("Gene") -->
<!-- # -->
<!-- # summary_table <- df %>% -->
<!-- #   mutate(row_total = rowSums(across(-Gene))) %>% -->
<!-- #   pivot_longer( -->
<!-- #     -c(Gene, row_total), names_to = "donor", values_to = "value" -->
<!-- #   ) %>% -->
<!-- #   group_by(donor) %>% -->
<!-- #   summarize( -->
<!-- #     total_genes  = sum(value == 1), -->
<!-- #     shared_genes = sum(value == 1 & row_total > 1), -->
<!-- #     unique_genes = total_genes - shared_genes, -->
<!-- #     prop_shared = shared_genes / total_genes, -->
<!-- #     .groups = "drop" -->
<!-- #   ) -->

<!-- reg <- per_donor_data[["logical_matrix"]] %>% -->
<!--   rownames_to_column("Gene") %>% -->
<!--   mutate(row_total = rowSums(across(-Gene))) %>% -->
<!--   pivot_longer( -->
<!--     -c(Gene, row_total), names_to = "donor", values_to = "value" -->
<!--   ) %>% -->
<!--   group_by(donor) %>% -->
<!--   summarize( -->
<!--     total_genes  = sum(value == 1), -->
<!--     shared_genes = sum(value == 1 & row_total > 1), -->
<!--     unique_genes = total_genes - shared_genes, -->
<!--     percent_shared = 100 * shared_genes / total_genes, -->
<!--     .groups = "drop" -->
<!--   ) %>% -->
<!--   mutate(criterion = "reg", .before = 1) -->

<!-- upreg <- per_donor_data[["logical_matrix_upreg"]] %>% -->
<!--   rownames_to_column("Gene") %>% -->
<!--   mutate(row_total = rowSums(across(-Gene))) %>% -->
<!--   pivot_longer( -->
<!--     -c(Gene, row_total), names_to = "donor", values_to = "value" -->
<!--   ) %>% -->
<!--   group_by(donor) %>% -->
<!--   summarize( -->
<!--     total_genes  = sum(value == 1), -->
<!--     shared_genes = sum(value == 1 & row_total > 1), -->
<!--     unique_genes = total_genes - shared_genes, -->
<!--     percent_shared = 100 * shared_genes / total_genes, -->
<!--     .groups = "drop" -->
<!--   ) %>% -->
<!--   mutate(criterion = "upreg", .before = 1) -->

<!-- downreg <- per_donor_data[["logical_matrix_downreg"]] %>% -->
<!--   rownames_to_column("Gene") %>% -->
<!--   mutate(row_total = rowSums(across(-Gene))) %>% -->
<!--   pivot_longer( -->
<!--     -c(Gene, row_total), names_to = "donor", values_to = "value" -->
<!--   ) %>% -->
<!--   group_by(donor) %>% -->
<!--   summarize( -->
<!--     total_genes  = sum(value == 1), -->
<!--     shared_genes = sum(value == 1 & row_total > 1), -->
<!--     unique_genes = total_genes - shared_genes, -->
<!--     percent_shared = 100 * shared_genes / total_genes, -->
<!--     .groups = "drop" -->
<!--   ) %>% -->
<!--   mutate(criterion = "downreg", .before = 1) -->

<!-- atLeastOneShared_summary <- rbind( -->
<!--   reg, -->
<!--   upreg, -->
<!--   downreg -->
<!-- ) -->

<!-- df <- atLeastOneShared_summary %>% -->
<!--   select(criterion, donor, `total genes` = total_genes, `percent shared` = percent_shared) %>% -->
<!--   pivot_longer(cols = c(`total genes`, `percent shared`), names_to = "stat", values_to = "value") %>% -->
<!--   mutate( -->
<!--     group = if_else(criterion == "reg", "all regulated genes", "split by direction"), -->
<!--     stat = factor(stat, levels = c("total genes", "percent shared")), -->
<!--     color = if_else(str_detect(donor, "A549"), "A549", "primary"), -->
<!--     label = if_else(str_detect(donor, "A549"), "A549", NA) -->
<!--   ) -->

<!-- print( -->

<!--   ggplot(df, -->
<!--          aes(x = criterion, y = value, color = color, label = label) -->
<!--   ) + -->
<!--     facet_grid(stat~group, scales = "free") + -->
<!--     stat_summary(geom = "bar", fun = "mean", width = 0.5, color = "black") + -->
<!--     # stat_summary(geom = "errorbar", fun.data = "mean_se", width = 0.3) + -->
<!--     geom_jitter(width = 0.05) + -->
<!--     theme_bw() + -->
<!--     theme( -->
<!--       strip.background = element_blank(), -->
<!--       panel.grid.minor = element_blank() -->
<!--     ) + -->
<!--     scale_y_continuous( -->
<!--       expand = expansion(mult = c(0, 0.25)), -->
<!--       breaks = scales::breaks_pretty(n = 5)) + -->
<!--     scale_color_manual(values = c("firebrick2", "black")) + -->
<!--     labs(y = NULL, x = NULL) + -->
<!--     guides(color = "none") + -->

<!--     geom_text_repel( -->
<!--       nudge_y = -5, nudge_x = -0.3 -->
<!--     ) -->

<!-- ) -->

<!-- cleanup_objects() -->
<!-- ``` -->

<!-- **Are primary cells more variable in IL1B-regulated gene expression than A549 cells?** Should be. Check with PCA below: -->

<!-- The PCA dispersion of HBEs is not statistically different from A549s, but ALIs is different from A549s. -->

<!-- ```{r} -->
<!-- reg_genes <- rownames(per_donor_data[["logical_matrix"]]) -->
<!-- upreg_genes <- rownames(per_donor_data[["logical_matrix_upreg"]]) -->
<!-- downreg_genes <- rownames(per_donor_data[["logical_matrix_downreg"]]) -->

<!-- pca_from_tpm <- function(.treatment, .genelist){ -->

<!--   mat <- tpm %>% -->
<!--     filter(treatment == .treatment, Gene %in% .genelist) %>% -->
<!--     mutate(name = paste(celltype, rep)) %>% -->
<!--     select(Gene, name, log2fold) %>% -->
<!--     pivot_wider(names_from = "name", values_from = "log2fold") %>% -->
<!--     column_to_rownames("Gene") -->

<!--   pca <- prcomp(t(mat), scale. = TRUE) -->

<!--   pca_df <- pca$x %>% -->
<!--     as.data.frame() %>% -->
<!--     rownames_to_column("name") %>% -->
<!--     select(name, PC1, PC2, PC3, PC4) %>% -->
<!--     separate(name, into = c("celltype", "rep"), sep = " ", remove = F) -->

<!--   centroids <- pca_df %>% -->
<!--     group_by(celltype) %>% -->
<!--     summarize(across(PC1:PC4, mean), .groups = "drop") -->


<!--   pca_df <- pca_df %>% -->
<!--     left_join(centroids, by = "celltype", suffix = c("", "_centroid")) %>% -->
<!--     rowwise() %>% -->
<!--     mutate( -->
<!--       dist_to_centroid = sqrt( -->
<!--         sum( -->
<!--           (c_across(PC1:PC4) - c_across(PC1_centroid:PC4_centroid))^2 -->
<!--         ) -->
<!--       ) -->
<!--     ) %>% -->
<!--     ungroup() -->

<!--   dispersion <- pca_df %>% -->
<!--     group_by(celltype) %>% -->
<!--     summarize( -->
<!--       mean_dist = mean(dist_to_centroid), -->
<!--     sd_dist   = sd(dist_to_centroid), -->
<!--     .groups = "drop" -->
<!--   ) -->

<!--   list( -->
<!--     "matrix" = mat, -->
<!--     "pca" = pca, -->
<!--     "summary" = pca_df, -->
<!--     "dispersion" = dispersion -->
<!--   ) -->

<!-- } -->

<!-- test <- pca_from_tpm("IL1B", reg_genes) -->

<!-- ggplot( -->
<!--   test$summary, -->
<!--   aes(x = PC1, y = PC2, shape = celltype) -->
<!-- ) + -->
<!--   geom_point() -->

<!-- ggplot( -->
<!--   test$summary, -->
<!--   aes(x = celltype, y = dist_to_centroid) -->
<!-- ) + -->
<!--   geom_boxplot() + -->
<!--   geom_point() -->

<!-- fit <- aov(dist_to_centroid ~ celltype, data = test$summary) -->
<!-- summary(fit) -->

<!-- tukey <- TukeyHSD(fit) -->
<!-- tukey -->

<!-- cleanup_objects() -->
<!-- ``` -->
```

