---
title: "A549-HBE-ALI comparison analysis"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

# for functional annotation
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)

# for set theory
library(ggVennDiagram)
library(ComplexUpset)
library(eulerr)

# heatmaps
library(ComplexHeatmap)
library(pheatmap)

# misc tools
library(patchwork)
library(ggrepel)
source("modules/helpers.R")

#tidyverse
library(tidyverse)
```

Repo at https://github.com/axg610/a549-hbe-ali-comparison

# **1. Alignment and DEA**

**Raw datasets:** A549, primary hBECs (submersion culture, HBE), and primary hBECs (air-liquid interface culture, ALI) treated with IL1B, glucocorticoid, or both at 6h. This is actually three experiments at different timepoints, so for now we use their shared timepoint (6 h post-treatment).

Alignment performed with kallisto version 0.46.1 and normalization/DEA performed with sleuth version ??? on the cluster. Everything documented in `kallisto-sleuth.md`. I used individual sleuth runs for each cell type, modeled as `~treatment`. Effect sizes ("log2fold") and FDR are from Wald test using unstimulated (NS) samples as baseline. I also modeled everything together under model `~treatment*condition` for interaction testing. Raw tpm values were also extracted.

```{r load data, cache=T}
dea <- read_tsv("data/raw/A549vsPrimary_univariateDEA.txt")
tpm <- read_tsv("data/raw/A549vsPrimary_tpm.txt")
multivar <- read_tsv("data/raw/A549vsPrimary_multivariateDEA.txt")
```

# **2. Story notes**

Given that IL1B-responsive gene expression patterns in A549s cells follow ... kinetics, carry ... functional signatures, we next want to compare the transcriptomic signatures of A549 cells and primary cells at 6 h (time of most regulation)... With the ultimate aim of understanding the extent to which A549 cells recapitulate donor responses to IL1B.

## 2.1. Small overlap between A549 cells and donors

**The A549 transcriptome overlaps poorly with the aggregate of all donors.** Venn diagrams were used to compare IL1B-responsive DEGs identified in A549 cell data and primary cell data (all donors pooled by cell type). Thresholds were set at log2fold >= 1 or -1 for induction amplitude and FDR < 0.05 for statistical significance. **(A)** Only ~17% of the IL1B-regulated transcriptome is shared with at least one aggregate primary cell type, increasing to ~33% for upregulated genes **(B)** and decreasing to ~4% for downregulated genes **(C)**. Trends remain stable when filtering with log2fold or FDR only **(D-I)**.

```{r old venns, fig.height=4, fig.width=4, out.width="33%", fig.show='hold', cache=T}
source("modules/fig-old-venns.R")
cleanup_objects()
```

**Heatmap from Venn A above.** A = A549, L = ALI, H = HBE. So "ALH" refers to the triple overlap, "A" refers to A549 only, etc.

```{r heatmap il1b responsive genes, fig.height=8, fig.width=4, out.width ="50%", fig.show="hold"}
source("modules/fig-vennA-heatmap.R")
cleanup_objects()
```


## 2.2. A549 cell replicates are relatively stable

*I feel like this section is here to justify collapsing A549 samples into one "mean." Don't really have to show it, but it's here in case reviewers ask. If this ever gets shown, reviewers might ask us to do the same thing in reps of primary cell donors.*

```{r calculate per-donor metrics, cache=T}
source("modules/per-donor-thresholding.R")
source("modules/compute-heavy-heatmaps.R")
cleanup_objects()
```

**A549 replicates are less variable than primary cell donors.** To understand variability within A549, ALI, and HBE samples, IL1B-regulated genes (abs(log2fold) >= 1) were identified in each replicate or donor. **(A)** A549 replicates show greater overlap in regulated genes than primary cells. **(B)** The union of all IL1B-regulated genes (14621) were visualized as a heatmap. Compared to primary cell donors, A549 replicates appear more coherent. Note "splotches" of gene expression unique to one primary cell donor, which do not appear in A549 cells. Accompanying basal tpm heatmap suggests these variable genes have lower baseline expression. 

**Variability between A549 replicates are related to basal expression.** **(C)** When ordering A549-regulated genes by SD across reps (stable on top, variable on bottom), variability appears inversely related to basal expression. **(D, E)** A549 genes were ordered by SD and divided into 10 bins. Bins with higher SD tend to have lower basal expression.


```{r venns and heatmaps of reps, fig.height=7, fig.width=6, out.width="50%", fig.show='hold'}
source("modules/fig-variability-due-to-basal.R")
cleanup_objects()
```


## 2.3. Donors are highly variable

**Primary cell donors are more variable than A549 cell replicates.** Counts and variability for regulated genes (abs(log2fold) > 1)  were assessed across donors. **(A-C)** A549 and primary cells have similar numbers of "regulated" genes, with slightly less upregulation in A549s. **(D)** For "regulated" genes in general, there is more variability in log2fold across replicates of A549s than pooled primary cell donors. **(E-F)** This relationship reverses when restricting the analysis to genes regulating in the same direction, suggesting that between-donor variability comes from failure to "agree" on the direction in which many genes regulate. **(G)** When visualizing the union of all regulated genes (14621) across replicates as a heatmap, primary cells show blocks of up- or down-regulation unique to each donor.

```{r plot per-donor figs, fig.height=5, fig.width=9}
source("modules/fig-per-donor-threshold-metrics.R")
cleanup_objects()
```

**Donors show more uniquely regulated genes than shared genes.** To understand individual variation, A549 cell replicates were averaged to represent one donor. Then overlaps between the A549 donor and all other primary donors were visualized as an UpSet plot. **(A)** Compared to genes regulated in only one donor (left side, first 11 bars), there are much fewer genes shared between donors (right side) *(do a manual count?)*. Splitting the visualization into **(B)** upregulated and **(C)** downregulated genes reveals that A549s have "normal" IL1B-upregulated signatures (middle-of-the-pack in terms of unique upreg genes), but abberant IL1B-downregulated signatures (has the most downregulated genes). Additionally, the 54 completely conserved genes are exclusively upregulated. This suggests that the IL1B-induced response is more core/conserved while the downregulated response is more context/individual-specific.

```{r plot upsets between donors, fig.height=8, fig.width=8}
source("modules/fig-per-donor-upset.R")
cleanup_objects()
```

**Transcriptomic overlaps between each donor and A549 cells are small.** To assess similarities between each primary cell donor and the A549 mean, regulated genes were compared between each primary cell donor and A549 cells. **(A)** Only a small percentage (<15%) of each donor's regulation pool is recapitulated by A549 cells. This is less of a problem for IL1B-induced genes **(B)** and more of a problem for IL1B-downregulated genes **(C)**. J = Jaccard index, n = total number of genes.

```{r plot pairwise venns vs a549, fig.width=8}
source("modules/fig-per-donor-pairwise-venns.R")
cleanup_objects()
```

Plotting each donor-to-A549 comparison, we see that overlaps tend to be bigger for upregulated genes and smaller for downregulated. T-test for $Jaccard_{\text{upreg}}$ vs $Jaccard_{\text{downreg}}$ (bottom right panel) returns P=0.0001169 (***).

```{r plot pairwise venns summary, fig.height=4, fig.width=4}
source("modules/fig-per-donor-pairwise-venns-summary.R")
cleanup_objects()
```

## 2.4. Majority of A549-induced genes are shared with donors

**More genes shared than unique**. For the A549 mean and all donors, genes were called regulated, upregulated, or downregulated if they met abs(log2fold) >= 1, log2fold >= 1, or log2fold <= -1 thresholds, respectively. Genes were then classified as unique to A549s, or shared between A549s and at least one primary cell donor. **(A)** For "regulated" genes, more are shared than unique (69% vs 31%). This does not factor in direction. **(B, C)** Similar trend for upregulated genes and less so for downregulated. Genes were only counted as shared if they regulated in the same direction.

This property is not unique to A549 cells. When taking each primary cell donor and comparing their unique genes vs those shared with at least one other donor (ALIs, HBEs, and A549), we get ~83% shared with 1+ other donor(s) and ~17% unique to that donor. Each point is one cell.

```{r fig.height=3, fig.width=4, out.width='50%', fig.show='hold'}
source("modules/fig-majority-shared-with-donors.R")
cleanup_objects()
```

## 2.5. Stable induction of inflammatory responses between donors

**Inflammatory programs are stably regulated across donors.** To understand which transcriptional programs are highly variable vs conserved across donors, the union of all IL1B-induced genes in each cell type were ranked by -SD across all donors. Thus, top ranking genes (green) are consistently/stably regulated by IL1B in all donors, while bottom ranking genes (purple) are highly variable between donors.

```{r run gene ontology on variable vs stable genes, cache=T}

go_results <- list()

go_results[["variable_vs_stable"]][["df"]] <- data.frame()

for(ct in c("A549", "ALI", "HBE")){
  
  # rank all "regulated" genes by SD between donors
  
  ranked <- tpm %>%
    filter(
      celltype == ct,
      treatment == "IL1B"
    ) %>%
    group_by(Gene) %>%
    filter(any(log2fold >= 1)) %>%
    summarize(sd = -sd(log2fold), .groups = "drop") %>%
    arrange(desc(sd)) %>%
    deframe()
  
  # ranked list of all genes
  all_genes <- names(ranked)
  
  # calculate and cut the top/bottom 10%
  n <- length(all_genes)
  cut <- floor(0.10 * n)
  top_genes <- all_genes[1:cut]
  bottom_genes <- all_genes[(n - cut + 1):n]
  
  # run and extract top results, enrichGO on biological process
  
  top_go_bp <- enrichGO(
    gene          = top_genes,
    universe      = all_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )
  
  top_go_result <- top_go_bp@result %>%
    mutate(
      celltype = ct,
      treatment = "IL1B",
      group = "top 10%",
      .before = 1
    )
  
  rownames(top_go_result) <- NULL
  
  # run and extract bottom results, enrichGO on biological process
  
  bottom_go_bp <- enrichGO(
    gene          = bottom_genes,
    universe      = all_genes,
    OrgDb         = org.Hs.eg.db,
    keyType       = "SYMBOL",
    ont           = "BP",
    pAdjustMethod = "BH",
    qvalueCutoff  = 0.05,
    readable      = TRUE
  )
  
  bottom_go_result <- bottom_go_bp@result %>%
    mutate(
      celltype = ct,
      treatment = "IL1B",
      group = "bottom 10%",
      .before = 1
    )
  
  rownames(bottom_go_result) <- NULL
  
  # save dataframes to main object
  
  go_results[["variable_vs_stable"]][["df"]] <- rbind(
    go_results[["variable_vs_stable"]][["df"]], 
    top_go_result
    )
  
  go_results[["variable_vs_stable"]][["df"]] <- rbind(
    go_results[["variable_vs_stable"]][["df"]], 
    bottom_go_result
  )
  
  # save genelists to main object
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_top10percentInduced")
  ]] <- top_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_bottom10percentInduced")
  ]] <- bottom_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_allInduced")
  ]] <- all_genes
  
  go_results[["variable_vs_stable"]][[
    paste0(ct, "_IL1B_allInduced_sd")
  ]] <- ranked
  
}

cleanup_objects()
```

```{r plot heatmap of sd rankings, fig.height=4, fig.width=4, out.width='50%', fig.show='hold'}

for(ct in c("ALI", "HBE")){
  
  # get matrix and genelists
  query <- paste0(ct, "_IL1B_allInduced")
  topquery <- paste0(ct, "_IL1B_top10percentInduced")
  bottomquery <- paste0(ct, "_IL1B_bottom10percentInduced")
  
  genelist <- go_results[["variable_vs_stable"]][[query]]
  top_genes <- go_results[["variable_vs_stable"]][[topquery]]
  bottom_genes <- go_results[["variable_vs_stable"]][[bottomquery]]
  
  mat <- tpm %>%
    filter(
      treatment == "IL1B",
      Gene %in% genelist,
      celltype == ct
    ) %>%
    select(Gene, rep, log2fold) %>%
    pivot_wider(values_from = "log2fold", names_from = "rep") %>%
    mutate(Gene = factor(Gene, levels = genelist)) %>%
    arrange(Gene) %>%
    column_to_rownames("Gene") %>%
    as.matrix()
  
  # set colors
  colors <- generic_l2f_heatmap_colors(mat)
  
  # assign groups by membership
  group <- case_when(
    rownames(mat) %in% top_genes ~ "top 10% -SD (most stable)",
    rownames(mat) %in% bottom_genes ~ "bottom 10% -SD (most variable)",
    TRUE ~ "middle"
  )
  
  group <- factor(
    group, 
    levels = c(
      "top 10% -SD (most stable)", 
      "middle", 
      "bottom 10% -SD (most variable)"
      )
    )
  
  # annotation
  row_ha <- rowAnnotation(
    group = group,
    col = list(
      group = c(
        "top 10% -SD (most stable)" = "aquamarine3",
        middle = "grey77",
        "bottom 10% -SD (most variable)" = "orchid3"
      )
    ),
    show_annotation_name = FALSE,
    border = TRUE
  )
  
  # draw hm
  hm <- Heatmap(
    mat,
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    show_column_dend = FALSE,
    show_row_names = FALSE,
    col = colors,
    name = "log2fold",
    column_title = ct,
    height = unit(3, "inches"),
    width = unit(1, "inches"),
    left_annotation = row_ha,
    border = TRUE
  )
  
  print(hm)
  
}

cleanup_objects()


```

The two extremes of these rankings (top and bottom 10% most variable, effectively the green and purple tips of the heatmap) were taken for gene ontology enrichment analysis (GO_BP_DIRECT), and then the top 10 terms in each case were plotted. Terms with less than 10 genes were excluded to prevent poorly defined programs from dominated the ranking. **(A)** In both ALI and HBE cells, donors show stable induction of inflammatory and apoptotic programs. **(B)** Both ALI and HBE cells are more variable in expression of metabolic programs. HBE cells showed some enrichment for inflammatory responses. However, none of these enrichments reached significance.

```{r plot variable vs stable GO, fig.width=5, fig.height=7, out.width='50%', fig.show='hold'}

# function to prepare dataframes

prep_go <- function(df, ct, grp) {
  
  df %>%
    filter(celltype == ct, group == grp, Count >= 10) %>%
    slice_max(FoldEnrichment, n = 10, with_ties = FALSE) %>%
    mutate(
      Description = str_wrap(str_trunc(Description, 60), 35),
      Description = factor(Description, levels = rev(Description))
    )
  
}



# function to generate the base plot

base_plot <- function(
    df, 
    title,
    size_limits = c(0, 30),
    x_limits = c(0, 10),
    fill_limits = c(0, 4)) {
  ggplot(
    df,
    aes(
      x = FoldEnrichment,
      y = Description,
      fill = -log10(p.adjust),
      size = Count
    )
  ) +
    geom_point(shape = 21) +
    scale_size(
      limits = size_limits, range = c(0, 10)
    ) +
    # scale_fill_gradient(
    #   low = "white", high = "black",
    #   limits = fill_limits
    # ) +
    viridis::scale_fill_viridis(
      option = "viridis",
      limits = fill_limits
    ) +
    scale_x_continuous(
      limits = x_limits,
      expand = expansion(mult = 0.15)
    ) +
    labs(
      x = "fold enrichment", 
      y = title,
      size = "count",
      fill = expression(log[10] * "FDR")
    ) +
    theme_minimal()
}


# ==============================================================================


# show top 10% most stable genes in HBEs

p1 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "HBE", 
  grp = "top 10%"
  ) %>%
  base_plot(
    title    = "HBE", 
    x_limits = c(2.3, 4.0)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )

# show top 10% most stable genes in ALIs
p2 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "ALI", 
  grp = "top 10%") %>%
  base_plot(
    title    = "ALI", 
    x_limits = c(2.3, 4.0)
  ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank()
  )


# ==============================================================================


# show top 10% most variable genes in HBEs

p3 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "HBE", 
  grp = "bottom 10%"
  ) %>%
  base_plot(
    title    = "HBE", 
    x_limits = c(1, 1.9)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank()
  )

p4 <- prep_go(
  df  = go_results[["variable_vs_stable"]][["df"]], 
  ct  = "ALI", 
  grp = "bottom 10%"
  ) %>%
  base_plot(
    title    = "ALI", 
    x_limits = c(1, 1.9)
    ) +
  theme(
    axis.title.y = element_text(size = 20),
    panel.grid.minor = element_blank()
  )


# ==============================================================================

# assemble plot

(p1 / p2) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "A. most stably induced functions")

(p3 / p4) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "B. most variably induced functions")

cleanup_objects()
```
